---
title: "Structure by habitat type"
output:
  html_document:
    df_print: paged
---

Load libraries:

```{r, echo=FALSE}
library(ape)
library(paran)
library(splits)
library(ggplot2)
library(dplyr)
```

Modify the plot function from splits (add a 2 so R don't get confused) | Se modifica la función de graficar del paquete splits (se agrega un 2 para que R no se confunda)
```{r}
plotCluster2 <- function(tr, lthresh, show.tip.label = FALSE, # tip labels to FALSE
show.node.label = FALSE, cex = 0.5, edge.width) {
numnod <- tr$Nnode
numtip <- length(tr$tip.label)
cdat <- array(1, 2 * numnod)
ndat <- array("", numnod)
bt <- -branching.times(tr)
nest.nodes <- function(tr, x, p = 0) {
numtip <- length(tr$tip.label)
nods <- array(NA, 0)
desc <- as.integer(tr$edge[, 2][tr$edge[, 1] == x])
if (desc[1] > numtip) {
nods <- c(nods, desc[1], nest.nodes(tr, desc[1]))
}
if (desc[2] > numtip) {
nods <- c(nods, desc[2], nest.nodes(tr, desc[2]))
}
if (length(nods) > 0) {
return(nods)
}
else {
return(NULL)
}
}
threshold.group <- function(mrcas) {
parent <- tr$edge[, 1]
child <- tr$edge[, 2]
thresh.group <- list()
thresh.time <- c()
mrcas <- mrcas + numtip
k <- 1
while (TRUE) {
times <- bt[mrcas - numtip]
thresh1.time <- min(times)
thresh1.node <- mrcas[which.min(times)]
mrcas <- mrcas[-which.min(times)]
if (length(mrcas) == 0) {
thresh.time <- c(thresh.time, thresh1.time)
thresh.group[[k]] <- thresh1.node
break
}
member <- thresh1.node
del <- c()
for (i in 1:length(mrcas)) {
par.nod <- parent[child == mrcas[i]]
t.par <- bt[par.nod - numtip]
if (t.par < thresh1.time) {
member <- c(member, mrcas[i])
del <- c(del, i)
}
}
thresh.time <- c(thresh.time, thresh1.time)
thresh.group[[k]] <- member
k <- k + 1
if (length(del) != 0) {
mrcas <- mrcas[-del]
}
if (length(mrcas) == 0) {
break
}
}
return(thresh.group)
}
group <- threshold.group(lthresh)
colors <- rainbow(length(group))
k <- 1
for (g in group) {
n.col.type <- rep(0, numnod)
for (j in 1:length(g)) {
n.col.type[g[j] - numtip] <- 2
n.col.type[nest.nodes(tr, g[j]) - numtip] <- 1
}
cdat[match(tr$edge[, 1], which((n.col.type == 1) |
(n.col.type == 2)) + numtip) > 0] <-
colors[k]
k <- k + 1
}
plot(tr, edge.color = cdat, use.edge.length = 1, show.node.label =
show.node.label,
show.tip.label = show.tip.label, no.margin = FALSE,
cex = cex, edge.width=edge.width)
}
```



Use the trees resulted from the GMYC analysis (see "otus_delimitation.Rmd"). First, get population names and habitat from sequences names. B corresponds to forest and Z to alpine grasslands. | Se usan los árboles que resultaron del análisis GMYC (explicado en "otus_delimitation.Rmd"). Primero, se obtienen los nombres de las poblaciones y el tipo de vegetación de los nombres de las secuencias. B corresponde a bosques, y Z a pastizal alpino.

```{r, fig.width=12, fig.height=7}
Ttree<-read.nexus("trees/CurculionidaeTtree")

# run gmyc
Tgmyc<-gmyc(Ttree, method="single", interval=c(0,10),quiet=F)

# change tips
x<-Tgmyc
nomtips<-substr((x$tr)$tip.label, 1,2)
nomtips #Population
samtips<-substr((x$tr)$tip.label, 3,3)
samtips #habitat
#Definir color por tipo de vegetación (samtips)
samtips<-as.factor(samtips)
levels(samtips)<-c("darkolivegreen3", "darkorange")
samtips

# plot tree according to habitat type
plotCluster2(x$tree, x$MRCA[[which.max(x$likelihood)]], edge.width=2)
tiplabels(pch=15, col=as.vector(samtips), adj=0, offset=0.51, cex=1.5)
tiplabels(text=nomtips, frame="none", col="black", cex=1, adj=0, offset = 0.018)

# plot Time vs likelihood
plot(x$threshold.time, x$likelihood,type = "l", xlab = "Time", ylab = "Likelihood", cex.lab=1.5, cex.axis=1.5)

# plot time vs number of entityes and threshold time: -0.03913536
plot(x$threshold.time, x$entity, type="s" , xlab = "Time", ylab = "N", cex.lab=1.5, cex.axis=1.5)
abline(v=-0.03913536, col="red")

```

Repeat with all trees w/o code:


Curculionidae

```{r, echo= FALSE, fig.width=12, fig.height=11}
Ttree<-read.nexus("trees/CurculionidaeTtree")

# run gmyc
Tgmyc<-gmyc(Ttree, method="single", interval=c(0,10),quiet=F)

# change tips
x<-Tgmyc
nomtips<-substr((x$tr)$tip.label, 1,2)
nomtips #Population
samtips<-substr((x$tr)$tip.label, 3,3)
samtips #habitat
#Definir color por tipo de vegetación (samtips)
samtips<-as.factor(samtips)
levels(samtips)<-c("darkolivegreen3", "darkorange")
samtips

## PLOT
layout(matrix(c(1,1,2,3), 2, 2, byrow = TRUE),
       heights=c(2,1))

# plot tree according to habitat type
par(mar=c(0, 4, 2.5, 4))
plotCluster2(x$tree, x$MRCA[[which.max(x$likelihood)]], edge.width=2)
tiplabels(pch=15, col=as.vector(samtips), adj=0, offset=0.51, cex=1.5)
tiplabels(text=nomtips, frame="none", col="black", cex=1, adj=0, offset = 0.018)
title(main="Curculionidae (Coleoptera) \na)",adj=0, cex.main=1.5)

# plot Time vs likelihood
par(mar=c(4, 4, 2, 4))
plot(x$threshold.time, x$likelihood,type = "l", xlab = "Time", ylab = "Likelihood", cex.lab=1.5, cex.axis=1.5, lwd=2)
title(main="b)",adj=0, cex.main=1.5)

# plot time vs number of entityes and threshold time
par(mar=c(4, 4, 2, 4))
plot(x$threshold.time, x$entity, type="s" , xlab = "Time", ylab = "N", cex.lab=1.5, cex.axis=1.5, lwd=2)
abline(v=-0.03913536, col="red", lwd=2)
title(main="c)",adj=0, cex.main=1.5)

```




Carabideae
```{r, echo=FALSE, results= "hide", fig.width=12, fig.height=12}
Ttree<-read.nexus("trees/CarabidaeTtree")

# run gmyc
Tgmyc<-gmyc(Ttree, method="single", interval=c(0,10),quiet=F)

# change tips
x<-Tgmyc
nomtips<-substr((x$tr)$tip.label, 1,2)
nomtips #Population
samtips<-substr((x$tr)$tip.label, 3,3)
samtips #habitat
#Definir color por tipo de vegetación (samtips)
samtips<-as.factor(samtips)
levels(samtips)<-c("darkolivegreen3", "darkorange")
samtips

## PLOT
layout(matrix(c(1,1,2,3), 2, 2, byrow = TRUE),
       heights=c(3,1.2))

# plot tree according to habitat type
par(mar=c(0, 4, 2.5, 4))
plotCluster2(x$tree, x$MRCA[[which.max(x$likelihood)]], edge.width=2)
tiplabels(pch=15, col=as.vector(samtips), adj=0, offset=0.51, cex=1.5)
tiplabels(text=nomtips, frame="none", col="black", cex=1, adj=0, offset = 0.018)
title("Carabidae (Coleoptera) \nb)",adj=0, cex.main=1.5)

# plot Time vs likelihood
par(mar=c(4, 4, 2, 4))
plot(x$threshold.time, x$likelihood,type = "l", xlab = "Time", ylab = "Likelihood", cex.lab=1.5, cex.axis=1.5, lwd=2)
title(main="b)",adj=0, cex.main=1.5)

# plot time vs number of entityes and threshold time
par(mar=c(4, 4, 2, 4))
plot(x$threshold.time, x$entity, type="s" , xlab = "Time", ylab = "N", cex.lab=1.5, cex.axis=1.5, lwd=2, ylim=c(1,20))
abline(v=-0.05120494, col="red", lwd=2)
title(main="c)",adj=0, cex.main=1.5)
```


Linyphiidae
```{r, echo=FALSE, results= "hide", fig.width=12, fig.height=32}
Ttree<-read.nexus("trees/LinyphiidaeTtree")

# run gmyc
Tgmyc<-gmyc(Ttree, method="single", interval=c(0,10),quiet=F)

# change tips
x<-Tgmyc
nomtips<-substr((x$tr)$tip.label, 1,2)
nomtips #Population
samtips<-substr((x$tr)$tip.label, 3,3)
samtips #habitat
#Definir color por tipo de vegetación (samtips)
samtips<-as.factor(samtips)
levels(samtips)<-c("darkolivegreen3", "darkorange")
samtips

## PLOT
layout(matrix(c(1,1,2,3), 2, 2, byrow = TRUE),
       heights=c(6,1))

# plot tree according to habitat type
par(mar=c(0, 4, 2.5, 4))
plotCluster2(x$tree, x$MRCA[[which.max(x$likelihood)]], edge.width=2)
tiplabels(pch=15, col=as.vector(samtips), adj=0, offset=0.51, cex=1.5)
tiplabels(text=nomtips, frame="none", col="black", cex=1, adj=0, offset = 0.018)
title("Linyphiidae (Aranae) \nc)",adj=0, cex.main=1.5)

# plot Time vs likelihood
par(mar=c(4, 4, 2, 4))
plot(x$threshold.time, x$likelihood,type = "l", xlab = "Time", ylab = "Likelihood", cex.lab=1.5, cex.axis=1.5, lwd=2)
title(main="b)",adj=0, cex.main=1.5)

# plot time vs number of entityes and threshold time
par(mar=c(4, 4, 2, 4))
plot(x$threshold.time, x$entity, type="s" , xlab = "Time", ylab = "N", cex.lab=1.5, cex.axis=1.5, lwd=2)
abline(v=-0.07269869, col="red", lwd=2)
title(main="c)",adj=0, cex.main=1.5)
```


Gnaphosidae
```{r, echo=FALSE, results= "hide", fig.width=12, fig.height=8}
Ttree<-read.nexus("trees/GnaphosidaeTtree")

# run gmyc
Tgmyc<-gmyc(Ttree, method="single", interval=c(0,10),quiet=F)

# change tips
x<-Tgmyc
nomtips<-substr((x$tr)$tip.label, 1,2)
nomtips #Population
samtips<-substr((x$tr)$tip.label, 3,3)
samtips #habitat
#Definir color por tipo de vegetación (samtips)
samtips<-as.factor(samtips)
levels(samtips)<-c("darkolivegreen3", "darkorange")
samtips

## PLOT
layout(matrix(c(1,1,2,3), 2, 2, byrow = TRUE))

# plot tree according to habitat type
par(mar=c(0, 4, 2.5, 4))
plotCluster2(x$tree, x$MRCA[[which.max(x$likelihood)]], edge.width=2)
tiplabels(pch=15, col=as.vector(samtips), adj=0, offset=0.51, cex=1.5)
tiplabels(text=nomtips, frame="none", col="black", cex=1, adj=0, offset = 0.018)
title("Gnaphosidae (Aranae) \nd)",adj=0, cex.main=1.5)

# plot Time vs likelihood
par(mar=c(4, 4, 2, 4))
plot(x$threshold.time, x$likelihood,type = "l", xlab = "Time", ylab = "Likelihood", cex.lab=1.5, cex.axis=1.5, lwd=2)
title(main="b)",adj=0, cex.main=1.5)

# plot time vs number of entityes and threshold time
par(mar=c(4, 4, 2, 4))
plot(x$threshold.time, x$entity, type="s" , xlab = "Time", ylab = "N", cex.lab=1.5, cex.axis=1.5, lwd=2)
abline(v=-0.02034695, col="red", lwd=2)
title(main="c)",adj=0, cex.main=1.5)
```




Bar plots of all OTU abundance by habitat type
```{r}
# read data
samples_otus<-read.delim("samples_OTUs.txt") %>%
              dplyr::arrange(., OTU) %>% # order by otu
              mutate(., OTU=as.factor(OTU)) # otu as character not number

levels(samples_otus$OTU)<-unique(samples_otus$OTU) # add levels so that it plots nice (cant use integer because there is not otu 9)

# plot
ggplot(samples_otus, aes(x=OTU, fill=vegetation_type)) +
    geom_bar() + theme_bw() +
    scale_fill_manual(values=c("darkolivegreen3", "darkorange"),
                      name="Vegetation type") +
    theme(legend.title=element_text(size=13),
          legend.text=element_text(size=12),
          axis.title.x = element_text(size=13),
          axis.title.y = element_text(size=13),
          axis.text.x = element_text(size=12),
          axis.text.y = element_text(size=12)) +
    scale_x_discrete(breaks=seq(0,40,5))


```

Bar plot by family:

```{r}
ggplot(samples_otus, aes(x=OTU, fill=vegetation_type)) +
    geom_bar() +  theme_bw() +
  facet_wrap(. ~ family, scales="free_x")  +
  scale_fill_manual(values=c("darkolivegreen3", "darkorange"),
                      name="Vegetation type") +
    theme(legend.title=element_text(size=13),
          legend.text=element_text(size=12),
          axis.title.x = element_text(size=13),
          axis.title.y = element_text(size=13),
          axis.text.x = element_text(size=10),
          axis.text.y = element_text(size=12),
          strip.text = element_text(size=12)) +
      scale_x_discrete(breaks=c(1:11, seq(12,37,3), 39:45))
```

