---
title: "OTUs delimitation using GMYC | Determinación de OTUs con gmyc"
output:
  html_document:
    df_print: paged
---

1. Load libraries | Se cargan las librerías necesarias

```{r}
#install.packages("splits", repos = "http://R-Forge.R-project.org")
library(paran)
library(splits)
library(ape)
library(MASS)
library(dplyr)

```


2. Load and name the ultrametric tree (previously built in BEAST) | Se carga el árbol ultramétrico (usé BEAST para calcularlos). Se tiene que nombrar al árbol como objeto nuevo.

```{r}
CurcuTtree<-read.nexus("trees/CurculionidaeTtree")
```

3. Run gmyc using CurcuTree | Se implementa gmyc con el nuevo árbol.

```{r}
CurcuTgmyc<-gmyc(CurcuTtree, method="single", interval=c(0,10),quiet=F)
```

4. Results | Resultados:
```{r}
CurcuTgmyc
summary(CurcuTgmyc)
plot(CurcuTgmyc)
```

To change the branches names | Para cambiar la fuente de los nombres de las ramas se modifica la función (plotCluster3)
```{r}
plotCluster3 <- function(tr, lthresh, show.tip.label = TRUE,
                         show.node.label = FALSE, cex = 0.5) {
  numnod <- tr$Nnode
  numtip <- length(tr$tip.label)
  cdat <- array(1, 2 * numnod)
  ndat <- array("", numnod)
  bt <- -branching.times(tr)
  nest.nodes <- function(tr, x, p = 0) {
    numtip <- length(tr$tip.label)
    nods <- array(NA, 0)
    desc <- as.integer(tr$edge[, 2][tr$edge[, 1] == x])
    if (desc[1] > numtip) {
      nods <- c(nods, desc[1], nest.nodes(tr, desc[1]))
    }
    if (desc[2] > numtip) {
      nods <- c(nods, desc[2], nest.nodes(tr, desc[2]))
    }
    if (length(nods) > 0) {
      return(nods)
    }
    else {
      return(NULL)
    }
  }
  threshold.group <- function(mrcas) {
    parent <- tr$edge[, 1]
    child <- tr$edge[, 2]
    thresh.group <- list()
    thresh.time <- c()
    mrcas <- mrcas + numtip
    k <- 1
    while (TRUE) {
      times <- bt[mrcas - numtip]
      thresh1.time <- min(times)
      thresh1.node <- mrcas[which.min(times)]
      mrcas <- mrcas[-which.min(times)]
      if (length(mrcas) == 0) {
        thresh.time <- c(thresh.time, thresh1.time)
        thresh.group[[k]] <- thresh1.node
        break
      }
      member <- thresh1.node
      del <- c()
      for (i in 1:length(mrcas)) {
        par.nod <- parent[child == mrcas[i]]
        t.par <- bt[par.nod - numtip]
        if (t.par < thresh1.time) {
          member <- c(member, mrcas[i])
          del <- c(del, i)
        }
      }
      thresh.time <- c(thresh.time, thresh1.time)
      thresh.group[[k]] <- member
      k <- k + 1
      if (length(del) != 0) {
        mrcas <- mrcas[-del]
      }
      if (length(mrcas) == 0) {
        break
      }
    }
    return(thresh.group)
  }
  group <- threshold.group(lthresh)
  colors <- rainbow(length(group))
  k <- 1
  for (g in group) {
    n.col.type <- rep(0, numnod)
    for (j in 1:length(g)) {
      n.col.type[g[j] - numtip] <- 2
      n.col.type[nest.nodes(tr, g[j]) - numtip] <- 1
    }
    cdat[match(tr$edge[, 1], which((n.col.type == 1) |
                                     (n.col.type == 2)) + numtip) > 0] <- colors[k]
    k <- k + 1
  }
  plot(tr, edge.color = cdat, use.edge.length = 1, show.node.label =
         show.node.label,
       show.tip.label = show.tip.label, no.margin = FALSE,
       cex = cex, font=1)
}


plotCluster2 <- function(tr, lthresh, show.tip.label = FALSE, # tip labels to FALSE
show.node.label = FALSE, cex = 0.5, edge.width) {
numnod <- tr$Nnode
numtip <- length(tr$tip.label)
cdat <- array(1, 2 * numnod)
ndat <- array("", numnod)
bt <- -branching.times(tr)
nest.nodes <- function(tr, x, p = 0) {
numtip <- length(tr$tip.label)
nods <- array(NA, 0)
desc <- as.integer(tr$edge[, 2][tr$edge[, 1] == x])
if (desc[1] > numtip) {
nods <- c(nods, desc[1], nest.nodes(tr, desc[1]))
}
if (desc[2] > numtip) {
nods <- c(nods, desc[2], nest.nodes(tr, desc[2]))
}
if (length(nods) > 0) {
return(nods)
}
else {
return(NULL)
}
}
threshold.group <- function(mrcas) {
parent <- tr$edge[, 1]
child <- tr$edge[, 2]
thresh.group <- list()
thresh.time <- c()
mrcas <- mrcas + numtip
k <- 1
while (TRUE) {
times <- bt[mrcas - numtip]
thresh1.time <- min(times)
thresh1.node <- mrcas[which.min(times)]
mrcas <- mrcas[-which.min(times)]
if (length(mrcas) == 0) {
thresh.time <- c(thresh.time, thresh1.time)
thresh.group[[k]] <- thresh1.node
break
}
member <- thresh1.node
del <- c()
for (i in 1:length(mrcas)) {
par.nod <- parent[child == mrcas[i]]
t.par <- bt[par.nod - numtip]
if (t.par < thresh1.time) {
member <- c(member, mrcas[i])
del <- c(del, i)
}
}
thresh.time <- c(thresh.time, thresh1.time)
thresh.group[[k]] <- member
k <- k + 1
if (length(del) != 0) {
mrcas <- mrcas[-del]
}
if (length(mrcas) == 0) {
break
}
}
return(thresh.group)
}
group <- threshold.group(lthresh)
colors <- rainbow(length(group))
k <- 1
for (g in group) {
n.col.type <- rep(0, numnod)
for (j in 1:length(g)) {
n.col.type[g[j] - numtip] <- 2
n.col.type[nest.nodes(tr, g[j]) - numtip] <- 1
}
cdat[match(tr$edge[, 1], which((n.col.type == 1) |
(n.col.type == 2)) + numtip) > 0] <-
colors[k]
k <- k + 1
}
plot(tr, edge.color = cdat, use.edge.length = 1, show.node.label =
show.node.label,
show.tip.label = show.tip.label, no.margin = FALSE,
cex = cex, edge.width=edge.width)
}

```

Plot with the new fuction plotCluster3 and add edges number to know wich one correspodns to each OTU| Graficarlo con la nueva función plotCluster3
```{r}
x<-CurcuTgmyc
plotCluster3(x$tree, x$MRCA[[which.max(x$likelihood)]])
edgelabels(frame="none", adj = c(0.5, -0.25), cex=0.5)
```

Repeat this with each taxonomic family using a loop | Se repite con cada familia taxonómica

```{r}

for (i in c("CarabidaeTtree", "LinyphiidaeTtree", "GnaphosidaeTtree")){

print(paste("Results for", i))

# Load and name the ultrametric tree

Ttree<-read.nexus(paste0("trees/",i))

# run gmyc
Tgmyc<-gmyc(Ttree, method="single", interval=c(0,10),quiet=F)

# results
Tgmyc
summary(Tgmyc)
plot(Tgmyc)

# Plot (with previously modified plotCluster3 fun to change branches names) 
x<-Tgmyc
plotCluster3(x$tree, x$MRCA[[which.max(x$likelihood)]])
edgelabels(frame="none", adj = c(0.5, -0.25), cex=0.8)
}

```

### Check OTUs are named correctly in the trees edges:

REad data with assiged OTUs, will be used to compare in the trees
```{r}
# read data
samples_otus<-read.delim("samples_OTUs.txt") %>%
              dplyr::arrange(., OTU) %>% # order by otu
              mutate(., OTU=as.factor(OTU)) # otu as character not number

levels(samples_otus$OTU)<-unique(samples_otus$OTU) # add levels so that it plots nice (cant use integer because there is not otu 9)
```


Curculionidae

```{r, results="hide", fig.width=12, fig.height=11}
Ttree<-read.nexus("trees/CurculionidaeTtree")

# run gmyc
Tgmyc<-gmyc(Ttree, method="single", interval=c(0,10),quiet=F)

x<-Tgmyc

# get which CROP OTU is each tip
nomtips<-(x$tr)$tip.label
nomtips<-as.data.frame(nomtips)

otu_tips<-inner_join(nomtips, samples_otus, by=c("nomtips"="sample_id"))


## PLOT
layout(matrix(c(1,1,2,3), 2, 2, byrow = TRUE),
       heights=c(2,1))

# plot tree according to habitat type
par(mar=c(0, 4, 2.5, 4))
plotCluster2(x$tree, x$MRCA[[which.max(x$likelihood)]], edge.width=2)
tiplabels(text=otu_tips$OTU, frame="none", col="black", cex=.60, adj=0, offset = 0.006)
tiplabels(text=(x$tr)$tip.label, frame="none", col="black", cex=.5, adj=0, offset= 0.016)
title(main="Curculionidae (Coleoptera) \na)",adj=0, cex.main=1.5)

# add OTU names
edgelabels(text=c("OTU 5", "OTU 4", "OTU 3", "OTU 2", "OTU 7", "OTU 6", "OTU 1"), edge=c(54, 51, 24, 15, 14, 13, 4), frame="none", adj = c(0.5, -0.25), cex=.7, date= .09)
```



Carabideae
```{r, echo=FALSE, results= "hide", fig.width=12, fig.height=12}
Ttree<-read.nexus("trees/CarabidaeTtree")

# run gmyc
Tgmyc<-gmyc(Ttree, method="single", interval=c(0,10),quiet=F)

# change tips
x<-Tgmyc

# get which CROP OTU is each tip
nomtips<-(x$tr)$tip.label
nomtips<-as.data.frame(nomtips)

otu_tips<-inner_join(nomtips, samples_otus, by=c("nomtips"="sample_id"))


## PLOT
layout(matrix(c(1,1,2,3), 2, 2, byrow = TRUE),
       heights=c(3,1.2))

# plot tree according to habitat type
par(mar=c(0, 4, 2.5, 4))
plotCluster2(x$tree, x$MRCA[[which.max(x$likelihood)]], edge.width=2)
tiplabels(text=otu_tips$OTU, frame="none", col="black", cex=.6, adj=0, offset = 0.006)
tiplabels(text=(x$tr)$tip.label, frame="none", col="black", cex=.5, adj=0, offset= 0.016)
title("Carabidae (Coleoptera) \na)",adj=0, cex.main=1.5)

# add OTU names
edgelabels(text=c("OTU 11", "OTU 10"), edge=c(38,1), frame="none", adj = c(0.5, -0.25), cex=.8, date= .2)
```



Linyphiidae. 

```{r, echo=FALSE, results= "hide", fig.width=13, fig.height=32}
Ttree<-read.nexus("trees/LinyphiidaeTtree")

# run gmyc
Tgmyc<-gmyc(Ttree, method="single", interval=c(0,10),quiet=F)

x<-Tgmyc

# get which CROP OTU is each tip
nomtips<-(x$tr)$tip.label
nomtips<-as.data.frame(nomtips)

otu_tips<-inner_join(nomtips, samples_otus, by=c("nomtips"="sample_id"))

####

# plot tree according to habitat type and otus name
par(mar=c(0, 4, 2.5, 4))
plotCluster2(x$tree, x$MRCA[[which.max(x$likelihood)]], edge.width=2)
tiplabels(text=otu_tips$OTU, frame="none", col="black", cex=.50, adj=0, offset = 0.006)
tiplabels(text=(x$tr)$tip.label, frame="none", col="black", cex=.4, adj=0, offset= 0.016)
title("Linyphiidae (Aranae) \na)",adj=0, cex.main=1.5)

# add OTU names
edgelabels(text=c("OTU 31", "OTU 28", "OTU 27",  "OTU 34", "OTU 26", "OTU 38", "OTU 25", "OTU 24", "OTU 23", "OTU 29", "OTU 22", "OTU 21", "OTU 20", "OTU 19", "OTU 18", "OTU 33", "OTU 18", "OTU 35", "OTU 17", "OTU 16", "OTU 32", "OTU 15", "OTU 14", "OTU 13", "OTU 36", "OTU 30", "OTU 37", "OTU 12"), 
           edge=c(256, 245, 236, 235, 220, 215, 203, 196, 192, 190, 187, 180, 159, 150, 130, 129, 122, 118, 103, 100, 98, 62, 37, 15, 13, 9, 8, 2),
           frame="none", adj = c(0.5, -0.25), cex=.6)
```



Gnaphosidae
```{r, echo=FALSE, results= "hide", fig.width=12, fig.height=8}
Ttree<-read.nexus("trees/GnaphosidaeTtree")

# run gmyc
Tgmyc<-gmyc(Ttree, method="single", interval=c(0,10),quiet=F)
x<-Tgmyc

# get which CROP OTU is each tip
nomtips<-(x$tr)$tip.label
nomtips<-as.data.frame(nomtips)

otu_tips<-left_join(nomtips, samples_otus, by=c("nomtips"="sample_id"))

## PLOT
layout(matrix(c(1,1,2,3), 2, 2, byrow = TRUE))

# plot tree according to habitat type
par(mar=c(0, 4, 2.5, 4))
plotCluster2(x$tree, x$MRCA[[which.max(x$likelihood)]], edge.width=2)
tiplabels(text=otu_tips$OTU, frame="none", col="black", cex=.7, adj=0, offset = 0.006)
tiplabels(text=(x$tr)$tip.label, frame="none", col="black", cex=.5, adj=0, offset= 0.016)
title("Gnaphosidae (Aranae) \na)",adj=0, cex.main=1.5)

# Plot OTUs name
edgelabels(text=c("OTU 45", "OTU 44", "OTU 43", "OTU 42", "OTU 41", "OTU 40", "OTU 39"), edge=c(38, 35, 32, 26, 23, 10, 3), frame="none", adj = c(0.5, -0.25), cex=.8, date= .1)
```




